---
output: github_document
bibliography: vignettes/references.bib
link-citations: yes
---

```{r echo=FALSE, results = 'asis'}
pkg <- "markovDP"

source("https://raw.githubusercontent.com/mhahsler/pkg_helpers/main/pkg_helpers.R")
pkg_title(pkg)
```

## Introduction
A Markov decision process (MDP) [@Bellman1957; @Howard1960] is a discrete-time 
stochastic control process. In each time step, an 
agent can perform actions which affect the system (i.e., may cause
the system state to change). The agent's goal 
is to maximize its expected
future rewards that depend on the sequence of system state and the
agent's actions in the future. Solving the MDP means finding 
the optimal (or at least a good) policy
that guides the agent's actions. 

The `markovDP` package provides the infrastructure to work with MDPs in R.
The focus is on convenience in formulating MDPs in multiple ways, the support of
sparse representations 
(using sparse matrices, lists and data.frames) and visualization of results. 
Some key components are implemented in
C++ to speed up computation.
It also provides to the following popular solving procedures:

* __Dynamic Programming__
  - Value Iteration [@Bellman1957]
  - Modified Policy Iteration [@Howard1960; @Puterman1978]
  - Prioritized Sweeping [@Moore1993]

* __Linear Programming__
  - Primal Formulation [@Manne1960]

* __Termporal Differencing__
  - Q-Learning [@Watkins1992]
  - Sarsa [@Rummery1994]
  - Expected Sarsa [@Sutton1998]

These implementations follow the description is [@Russell2020] and 
[@Sutton1998].


```{r echo=FALSE, results = 'asis'}
pkg_citation(pkg, 1)
pkg_install(pkg, CRAN = FALSE)
```

## Usage

Solving the simple maze from [@Russell2020].
```{r problem}
library("markovDP")
data("Maze")
Maze
gridworld_plot(Maze)
```

```{r solve}
sol <- solve_MDP(model = Maze)
sol
```

Display the value function.

```{r value_function}
plot_value_function(sol)
```

```{r gridworld_plot}
gridworld_plot(sol)
```


## Acknowledgments

Development of this package was supported in part by 
National Institute of Standards and Technology (NIST) under grant number 
[60NANB17D180](https://www.nist.gov/ctl/pscr/safe-net-integrated-connected-vehicle-computing-platform).

## References
